\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing economic inequality with tax data - Switzerland from 1945 to 2010}

### 2014-07-02 paper/subdocuments/methods.Rnw
completed relative distribution part in methods.Rnw

In the last section we described the advantages and drawback of taxdata discussing five aspects, which we regard as crucial concering the assessment of inequality. To get a feeling of the importance of these aspects, we exploit the FTA-tax data as far as possible and perform several insightful calculations adressing four of the five mentioned aspects. No further investigation is possible regarding aspect (1) concepts on measuring economic ressource. This is a point which cannot be further addresses with FTA tax data. But for the other four thematic areas, we can provide deeper insight. In general our main strategy is to apply different possible concepts within one conceptional area (income, units, measurement, population coverage) while holding other conceptual differences constant. With this strategy we want to show, the sensitiveness the assessement of inequality is especially if one looks at time trens.  In this section we  describe which methods we use to produce the results in the next section.

@@

\textbf{Incomplete coverage of the population (left censored data.)} What can be done about the not-taxed? \citet{dell_income_2007} impute for non-fillers the 20 percentage of the annual average income. This flattens the distribution on the left side, which is not a problem if you are interested in the top income shares, but it would surly affect overall measures of inequality. Furthermore the authors calculate the proportion of non-fillers by estimating the total of tax units out of the population records. \\

\textbf{changes in taxation system  (switch from annual to biannual taxation)} In the mid-1990s a fundamental change in the Swiss tax system took place by switching form the two-years based praenumerando taxation to the one-year based postnumerando taxation. This change was enacted with a transitional period of several years, during which each canton could choose when to adopt the new system.  This is why during the transitional period from 1995 to 2003 there is no uniform tax data published on the Swiss level but only data on the cantonal level \citep[8f]{foellmi_volatile_2013}. \\
%es wird erwartet, dass der Wechsel Ungleichheitsmasse beeinflussen. Yearly fluctuations are dampened, when income is measured on a two-yearly basis.

\textbf{Estimating percentiles from bracket income tabulation} Pareto interpolation \\

\textbf{Missing of mean-tested benefits as part of the income} -> imputation with recommendation for minimum level for basic needs defined by the SKOS.\\
%is never mentioned as a problem, but it seems to me a better way to approach the non-taxed issue, than dell way (20 % of average income)
% Beim Imputieren m?sste man auch die Zahl der freiwilligen Nichtbeziehenden ber?cksichtigen.

\textbf{deductions} \citet[477]{dell_income_2007}:" we can check with statistics for 1971-72 (as well as later years) presented both by size of income before deductions and income after deductions that adding back deductions does not introduce any significant error in our estimates."
\citet[5]{schaltegger_evolution_2011}: ``..., information on [...] deductions is provided in the tax statistics, thus, we could add the personal deductions to the income data to obtain a consistent series over time''. Können wir das auch? Zumindest für gewisse Zeiträume? Das wäre noch gut. \\

Studies on income try to focus on the disposable income, which subtracts certain expenditures from the primary income. Deductions reflect somehow compulsory expenditures and thus taxable income can be seen as a sort of pseudo disposable income. On the other hand deductions can affect the distribution. There are recent studies about the correlation of progressivity and deductions in Switzerland, which examines if deductions have a ``perverse redistribution'' effect by redistributing income from the lower middle class to the upper middle class (see \citealt{peters_steuerabzuge:_2011} and \citet{Interpellation Barbara Gysel (2009) FEHLT. das interpellation barbara sieht falsch aus}.

>>>>

The RD-framework is based on the concept of a ``relative distribution'', a transformation of the data from two distributions into a single distribution that contains all of the information necessary for scale-invariant comparison. This allows the make distribution differences ``visual'' in an elegant way and it is also a base for summary statistics, which are more sensitive to detailed theoretical hypotheses in contrast to other measures like the gini-coefficient or top income-shares, which inform either about the whole population or only about the top part of the distribution.\\

% Probability Density function as a base for the RD
The goal of RD is to study the differences between two distributions. A common example could be the income distribution for man and women. Subject of the comparison can also be two distributions describing the same population, but stemming from two different data sources like survey or tax data or even comparison of the same source/population but for different time points, like the income distribution out of tax data for Switzerland today compared to an earlier time point. To describe what we gone compare, we start with defining the two distributions. On represents the reference population (\Y_{0}) and the other the comparison population \Y. Y represents our measure of interest (income). A first visual approach is to compare the two probability density functions (PDF). The PDF is a function (fy) which describes the distribution of probability over the outcome set and is defined for all possible values of y. Also this function integrates to 1, which means that the sum of all probabilities over all possible values is 100\%. Out of the comparison of the PDF, it is possible to see, which values of x are more and less probable. This allows allready to spot distributional differences over the whole scale of Y visualy. \\

The PDF can be characterized by its cumulative distribution function (CDF). The CDF can be formulated as F(y), which represents the cumulative probability at point x, which is the probability that a randomly chosen value is less than or equal to y. The relative distribution of \Y to \Y_{0} is then defined as

\begin{equation}
R=F_{0}(Y)
\end{equation}

R is obtained from \Y by transforming it by the CDF for \Y_{0}, \F_{0}. R measures therefore the relative rank of Y compared to Yo.

% Relative Distribution

\begin{equation}
g(r)=\frac{f(F_{0}^{-1}(r))}{^{f_{0}(F_{0}^{-1}(r))}}
\end{equation}

We can calculate the Probability Density Function (g(r)) of RD, where \r represents the proportion of values and F_{0}^{-1}(r) is the inverse cumulative distribution function, also called the quantile function. g(r) can be interpreted as a density Ratio, which is defined as the ratio of these two quantities of every percentile of the reference distribution [0,1]. With a complete overlap of both distributions the Probability Density Function of the RD is 1 at every point of the PDF. On the other hand, values higher than 1 represent higher probabilities in the comparison distribution than in the references distribution at this specific point and values lower than 1 respectively represent lower probabilities. It is a proper PDF in the sense that it integrates to 1 over the unit interval.\\

% Median and shape differences
What we get through the above transformation of two distributions is the overall relative probability density. But Differences between distributions can be divided into two basic components: changes in location and changes in shape. If the comparative distribution is a simple location-shifted version of the reference distribution, then the difference between the two distributions can be parsimoniously summarized by this shift. Differences that remain after a location adjustment are differences in ``shape'' (scale, skew and other distributional characteristics). When both types of shifts are operating, or when factors other than scale are changing in the shape component, we need a way to separate out the various effects. If we want to identify the effect of a location shift and separate it from other changes in the distribution, it is necessary to specify what scale this shift operates on. It is possible to adjust distribution by any measure of central tendency. We choose here a median location adjustment because the median is a natural, robust and scale invariant unit of measurement. Because our interest lies in analyzing distributional differences concerning the degree of inequality, we will focus in the results section on shape differences and look therefore at the relative distribution after the distributions are adjusted for location differences.

% RD based Summary Meausres
Distributional polarization is of particular interest in the study of inequality. However, common inequality indicator (for example Gini or Theil’s index) are not designed to distinguish between growth in the upper and lower tails. Even if the measures register increasing inequality over time, one cannot distinguish a polarization of the distribution (increases in both tails) from upgrading (increases in the upper tail), downgrading (increases in lower tail). The polarization index plays the same role as the difference in Gini-indices, but it can be decomposed to compare the growth in the upper and lower tails. Because it is based on the relative distribution it provides a simple link between what is observed in the graphical display and what is measured by the numerical summary. \\

The median relative polarization index (MRP) is then the mean absolute deviation around the median of the location-adjusted relative distribution, scaled to produce an index that varies between -1 and 1. Given the scaling, a value of zero represents no differences in distributional shape; positive values represent more polarization (increases in the tails of the distribution); and negative values represent less polarization (convergence towards the center of the distribution). The measure catches only differences in distributional shape (not location). And it has several interesting features. MRP can be interpreted in terms of a proportional shift of mass in the distribution from more central to less central values. A value of 0.1, for example, is equivalent to a 10\% population shift from the center of the distribution to the upper and lower quartiles and the MRP is decomposable along the scale of y. This makes it possible to compare the contribution of each section of the distribution to the overall polarization. A natural decomposition is the contributions made by components above and below the median (of g(r)).

@@

%\textbf{Incomplete coverage of the population (left censored data.)} What can be done about the not-taxed? \citet{dell_income_2007} impute for non-fillers the 20 percentage of the annual average income. This flattens the distribution on the left side, which is not a problem if you are interested in the top income shares, but it would surly affect overall measures of inequality. Furthermore the authors calculate the proportion of non-fillers by estimating the total of tax units out of the population records. \\

%\textbf{changes in taxation system  (switch from annual to biannual taxation)} In the mid-1990s a fundamental change in the Swiss tax system took place by switching form the two-years based praenumerando taxation to the one-year based postnumerando taxation. This change was enacted with a transitional period of several years, during which each canton could choose when to adopt the new system.  This is why during the transitional period from 1995 to 2003 there is no uniform tax data published on the Swiss level but only data on the cantonal level \citep[8f]{foellmi_volatile_2013}. \\
%es wird erwartet, dass der Wechsel Ungleichheitsmasse beeinflussen. Yearly fluctuations are dampened, when income is measured on a two-yearly basis.

%\textbf{Estimating percentiles from bracket income tabulation} Pareto interpolation \\

%\textbf{Missing of mean-tested benefits as part of the income} -> imputation with recommendation for minimum level for basic needs defined by the SKOS.\\
%is never mentioned as a problem, but it seems to me a better way to approach the non-taxed issue, than dell way (20 % of average income)
% Beim Imputieren m?sste man auch die Zahl der freiwilligen Nichtbeziehenden ber?cksichtigen.

%\textbf{deductions} \citet[477]{dell_income_2007}:" we can check with statistics for 1971-72 (as well as later years) presented both by size of income before deductions and income after deductions that adding back deductions does not introduce any significant error in our estimates."
%\citet[5]{schaltegger_evolution_2011}: ``..., information on [...] deductions is provided in the tax statistics, thus, we could add the personal deductions to the income data to obtain a consistent series over time''. Können wir das auch? Zumindest für gewisse Zeiträume? Das wäre noch gut. \\

%Studies on income try to focus on the disposable income, which subtracts certain expenditures from the primary income. Deductions reflect somehow compulsory expenditures and thus taxable income can be seen as a sort of pseudo disposable income. On the other hand deductions can affect the distribution. There are recent studies about the correlation of progressivity and deductions in Switzerland, which examines if deductions have a ``perverse redistribution'' effect by redistributing income from the lower middle class to the upper middle class (see \citealt{peters_steuerabzuge:_2011} and \citet{Interpellation Barbara Gysel (2009) FEHLT. das interpellation barbara sieht falsch aus}.
